Imperfect slope measurements drive overestimation in geometric cone model of waterbody depth

### 2021-11-10 manuscript/manuscript.Rmd
stash progress

Given the limited prediction accuracy of prior depth prediction efforts (± 6-7 m, Heathcote et al. 2015, Oliver et al. 2016, Sobek et al. 2011), a major focus has been on improving accuracy using strategies such as employing more diverse covariates (Oliver et al. 2016), varying waterbody buffer sizes (Heathcote et al. 2015), or estimating hidden groupings (e.g. fitting different models for distinct size classes) among waterbodies (Cael et al. 2017, Sobek et al. 2011). Unfortunately, the predictive accuracy of these efforts has been limited (± 6-7 m).

One intuitive approach for predicting waterbody depth involves using a geometric model that assumes waterbody basins correspond to an idealized shape such as a cone, bowl, or an elliptic sinusoid (Getirana et al. 2018, Hollister et al. 2011, Neumann 1959, Yigzaw et al. 2018). All such geometric models for waterbody depth prediction involve implicit assumptions about the terms of geometric formulae. In the simplest case, where waterbody basins are treated as cones (Eq. \ref{eq1}, Fig. 1), two assumptions are required to make depth predictions for all waterbodies: 1) that nearshore land slope is a representative proxy for in-lake slope and 2) that the distance to the center of the waterbody is a representative proxy for the distance to the deepest point of the waterbody (Fig. 1). This cone model imposes the following fixed (i.e. geometric) relationship between slope and horizontal distance:

@@

where the product of slope and horizontal distance yields an exact geometric depth estimate (depthgeometric). Cone models of waterbody basins have been used extensively to estimate hypsography in waterbodies where there is no knowledge of volume or mean depth (Read et al. 2014, Winslow et al. 2017).

@@

The assumptions of the cone model (as well as other geometric models) can be tested by comparing proxy measures of waterbody geometry against corresponding “true” (i.e. in-lake) values derived from bathymetric maps and by evaluating how waterbody cross-section shapes differ from that of an idealized cone (Johansson et al. 2007). For instance, waterbody cross-section shapes have been shown to vary from narrow "convex" forms to outstretched "concave" forms (Hakanson 1977). Because tests of geometric model assumptions require bathymetric map data, which is only available for a small fraction of waterbodies (including about 15% of all waterbodies in our study footprint), existing evidence may not be applicable to all waterbodies. The few studies that have tested these assumptions have been limited to individual studies of very large (> 500 ha) waterbodies or studies on small numbers (< 100) of waterbodies (Johansson et al. 2007). Studies focused specifically on reservoirs (as opposed to the more typical case where reservoirs and natural lakes are combined), have been even more restricted to that of large waterbodies > 1000 ha (Lehner et al. 2011, Messager et al. 2016).

As a result of this limited testing, we lack knowledge on both the predictive performance of geometric models, the effect of proxies on depth prediction, and whether depth predictions are more sensitive to measurement errors in the horizontal dimension (i.e. distance to the deepest point of the waterbody) or measurement errors in the vertical dimension (i.e. in-lake slope). Additionally, it is unclear whether model prediction error is related to differences in waterbody type such those with different cross-section shapes (concave versus convex) or those classified as reservoirs versus natural lakes. To address these knowledge gaps, we asked three research questions: (1) How representative is nearshore land slope of in-lake slope; and how representative is the distance to the center of a waterbody compared to the distance to the deepest point of a waterbody? (2) How does the use of proxies for waterbody geometry affect waterbody depth prediction error? (3) How does waterbody cross-section shape (i.e. concave versus convex) and waterbody type (i.e. natural lake vs reservoir) affect depth prediction error? To answer these questions, we extracted maximum depth (hereafter referred to as “observed maximum depth”), in-lake slope, cross-section shape (i.e., concave versus convex), and distance to the deepest point, of approximately 5,000 waterbodies from bathymetric map data. We supplemented these geometry measures with data classifying waterbodies as reservoirs or natural lakes. We used this data to compute geometric depth estimates (Equation 1) and prediction "offsets" to these estimates using the random forest algorithm (Equation 3). Such offsets are model quantities which minimize differences between observed and predicted depth as a function of covariates. In our case, covariates included a variety of waterbody, watershed, and hydrologic subbasin measures that are available for all waterbodies (Table 1).

By definition, the distance proxy (distance to the center of the waterbody) must always be greater or equal to the true distance value (distance to the deepest point of the waterbody). Therefore, we expect that the use of this proxy will lead to overestimation of waterbody depth (Fig. 1). Furthermore, we expect to see greater overestimation error in reservoirs as compared to natural lakes because many reservoirs are known to be drowned river valleys where the deepest point is close to the edge at the end of the reservoir (i.e. next to the dam) rather than in the center of the reservoir (Lanza and Silvey 1985). In a similar fashion, we expect to see overestimation error associated with using a nearshore land slope proxy in waterbodies with differing cross-section shape such that the depth of bowl-shaped (i.e. concave) waterbodies will be overpredicted whereas the depth of V-shaped (i.e. convex) waterbodies will be underpredicted (Fig. S1). Finally, we expect that depth predictions themselves will be strongly related to waterbody area and hydrologic subbasin variables as these measures have been influential in prior studies (Oliver et al. 2016).

@@

## Methods
### Data description
We compiled bathymetry data on approximately 5,000 waterbodies in the Northeastern and Midwestern United States from nine official state databases (Fig. 2). These represent approximately 15% of all waterbodies in the States included in our study and are a diverse cross section in terms of their characteristics, surface areas, and span a wide geographic extent including glaciated and non-glaciated regions (Table 1). Thus, they can be considered representative of the entire population of lakes in our study extent.

@@

The original data came in a variety of formats including pre-interpolated rasters (Minnesota), contour lines (Nebraska, Michigan, Massachusetts, Kansas, Iowa), contour polygons (New Hampshire, Connecticut), or point depth soundings (Maine). For the Minnesota data, we simply clipped the raster for each waterbody to its outline. For data from the remaining States, we processed each waterbody by converting its original representation to a point layer (if necessary), rasterizing these points, and creating an interpolated bathymetry “surface” using a simple moving window average in the raster R package (Hijmans 2019). The size of the moving window was adjusted iteratively to ensure that each bathymetry raster contained no missing data.

All waterbody bathymetry was specifically calculated relative to high-resolution (1:24,000 scale) National Hydrography Dataset (USGS 2019) waterbodies such that source data and bathymetry surface outputs were clipped to the area of each waterbody polygon. We restricted the waterbodies in our study to those with an area of at least 4 ha and a maximum depth of at least 0.3 m (1 ft.). The purpose of these restrictions was to ensure that waterbodies had enough contours (or points, or polygons) to generate adequately smooth interpolations with which to calculate in-lake geometry metrics.

@@

We categorized waterbodies based on their cross-section shape and reservoir class (e.g. natural lake, reservoir). For cross-section shape, we categorized waterbodies as either convex or concave following the method of Hakanson (1977) by computing normalized waterbody depth-area relationships (i.e. hypsographic curves) and assigning class membership based on whether the midpoint of a waterbody’s curve falls above or below that of a simple straight-sided cone (Fig. S2).

We further categorized waterbodies using the output of a deep convolutional neural network model trained on satellite images labelled according to whether there was visual evidence of a water control structure significantly impacting flow (Polus et. al 2021). This model had an overall validation accuracy of 81% and produced a probability for each waterbody as to whether it is a reservoir or a natural lake. For our purposes, we set a conservative classification probability threshold of 0.75 to determine whether a waterbody would be considered a reservoir. For example, if the Polus et al. (2021) dataset classified a particular waterbody as a reservoir with a probability of 0.74 we categorized it as a natural lake but if the probability was greater than 0.75 we categorized it as a reservoir. Note that our reservoir classification defines reservoirs as any permanent waterbody that has a water control structure likely to significantly impact flow or pool water. It makes no distinction between different dam types, heights, or uses/purposes because the Polus et al. (2021) dataset is only based on visual interpretation of waterbody images (via deep convolutional neural network models). However, the Polus et al. (2021) dataset is unique in that it provides data using a standardized approach at broad spatial extents for waterbodies > 4 ha.

@@

where sinuous waterbodies have larger values of shoreline development and circular waterbodies have smaller values of shoreline development. Watershed to waterbody area ratio is an approximation of water residence time and is defined as watershed area divided by waterbody area (Timms 2009).

### Proxy evaluation
We conducted a qualitative assessment of whether or not proxy measures of waterbody geometry (e.g.  nearshore land slope, distance to the center of the waterbody) are representative of their true values (e.g. in-lake slope, distance to the deepest point of the waterbody) by visual inspection (i.e. plotting each proxy measure against its corresponding true value) and by computing coefficients of determination (R2). We further tested proxy measures by examining their effect on waterbody depth prediction error. Our approach involved several steps. In the first step, we computed a geometric estimate of waterbody depth using only geometry information (depthgeometric, Equation 1). In the second step, we fit a random forest model to predict observed (i.e. true) depth as a function of geometric depth along with several covariates available for all waterbodies (Table 1). The purpose of this random forest “offset” modeling was to more rigorously test our expectations regarding prediction error among different formulations of depthgeometric and among different waterbody types. Each of these steps were executed iteratively for each combination of true and proxy values of slope and distance (Table 2). We conducted additional sensitivity analysis to examine possible interactions between different proxy measures of waterbody geometry and different subsets of the entire dataset where model data was restricted (i.e. “filtered”) to include only reservoirs, only natural lakes, only convex waterbodies, or only concave waterbodies (Table 2).

## Model description
### Geometric model 
We used a geometric model of waterbodies where basins are treated as cones with a fixed relationship between slope and distance (Equation 1). One reason that we used the cone model is that, unlike other idealized shapes, it does not require any knowledge of waterbody volume or mean depth. Note that Equation 1 is a geometric formula and has no intercept or coefficients and it produces an exact depth value given true values of slope and distance. To use this model to predict the depth of all waterbodies, there is a necessary assumption that proxy slope and distance measures, which are available for all waterbodies, are representative of true slope and distance (Fig. 1).

### Random forest models
Prior studies using geometric models to predict waterbody depth include a statistical or machine learning model “layer” or “offset” to boost predictive accuracy (Hollister et al. 2011, Yigzaw et al. 2018). This procedure involves fitting a statistical or machine learning model to the residuals of an initial geometric model. For our purposes, such offset modeling enabled us to test our expectations that prediction error would be different among different formulations of depthgeometric and among different waterbody types. It also facilitated direct comparison against prior models of waterbody depth including those that are non-geometric. We generated an “offset” to geometric depth (sensu Hollister et al. 2011) using the random forest algorithm and the ranger R package (Wright and Ziegler 2017) to predict observed maximum depth as a function of covariates including geometric maximum depth (from Equation 1) along with the waterbody elevation, area, perimeter, and ratio/index measures listed in Table 2:

@@

We evaluated the relative “importance” of individual covariates by comparing model performance between 1) models where a given covariate was left untouched versus 2) models where a given covariate was permuted randomly (Prasad et al. 2006, Wright and Ziegler 2017). Neither cross-section shape nor reservoir class was used as a covariate in any random forest models. Random forest training and test data was stratified on shape and reservoir class to match that of the overall waterbody population. We used the random forest algorithm because it makes no assumptions about the distribution of model residuals, allows for non-linearity, and is insensitive to interactions (i.e. multicollinearity) among covariates (Prasad et al. 2006).

### Model comparisons
We tested model sensitivity to slope and distance proxies by generating multiple “geometric maximum depth” estimates from 3 different model runs using each of the possible metric combinations for Equation 1 (true slope - proxy distance, proxy slope - true distance, proxy slope - proxy distance). Prior to entry into Equation 1, we standardized proxy distances to have the same numeric range as their true counterpart. The purpose of this standardization was to prevent waterbodies with extremely long proxy distances from having an outsized impact on model evaluation metrics. In addition to comparing among model runs using different metric combinations, we compared among sets of model runs where slope and distance measures were calculated using different sets of calculation techniques (Table S2).

### Model evaluations
We evaluated model fit and prediction error using root-mean-square error (RMSE), mean absolute percent error (MAPE), and coefficient of determination (R2) metrics on a holdout set (i.e. a data subset not used for model training) containing 25% of all waterbodies. We evaluated the residuals of each model relative to waterbody cross-section shape and reservoir classes to determine whether depth is consistently over or under predicted for some waterbody types relative to others.

@@

Although proxy distance to waterbody center was often larger in magnitude compared to the true distance to the deepest point of waterbodies’ (rather than being identical), they were strongly related (R2 = 0.8). Note that the coefficient of determination for this relationship is not strictly correct given that distance to waterbody center is an upper bound on distance to the deepest point of waterbodies. In contrast to distance metrics, proxy nearshore land slope and true in-lake slope were more weakly related (R2 = 0.17). For slope measures, most waterbodies had higher magnitude (i.e. steeper) nearshore land slope compared to true in-lake slope (Fig. 3). Taken together, these results suggest that proxy distance to the center of waterbodies is representative of true distance to the deepest point of waterbodies whereas proxy nearshore land slope is not representative of true in-lake slope. The strong relationship between distance to the center of waterbodies and distance to the deepest point means that it is possible to convert between the two measures in subsequent analyses (See best-fit equations in Fig. 3).

@@

The use of proxy nearshore land slope had a larger effect on model fit and prediction error than the use of proxy distance to waterbody center (Table 2). More specifically, the true slope (in-lake slope) - proxy distance (distance to the center of the waterbody) model had a better fit (R2 = 0.73) and lower prediction error (RMSE = 4.8m, MAPE = 27%) compared to the proxy slope - true distance model (R2 = 0.31, RMSE = 7.3m, MAPE = 64%). The fit of the proxy slope -proxy distance model (R2 = 0.36, RMSE = 7.1m, MAPE = 61%) was very similar to the proxy slope – true distance model. Predicted depth values for this model were generally underestimates relative to measured depth values (Fig. S9).

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.45\textwidth,keepaspectratio]{../figures/02_depth_model_importance-1.pdf}
  \end{center}
  \caption{Importance plot for random forest variables showing increase in mean square error. Higher values indicate greater importance to model predictions. See Equation 1 for a definition of geometric max depth. HUC4 ID is a 'dummy' variable of geographic (hydrologic subbasin) location.}\label{fig5}
\end{figure}

@@

### Representativeness of proxy measures of waterbody geometry
In comparing among waterbody geometry measures, our analysis suggests that proxy distance to waterbody center is representative of true distance to the deepest point of the waterbodies, but that proxy nearshore land slope is not representative of true in-lake slope. A simple indication of this non-representativeness is that proxy nearshore land slope was often (in > 74% of cases) steeper than true in-lake slope. This finding is consistent with Heathcote et al. (2015) whose results suggest that in-lake slopes are shallower compared to the surrounding land. Furthermore, the fact that in-lake slopes were shallower compared to the surrounding land even after controlling for differences in area (Fig. S4) is consistent with the idea of topographic scaling (i.e. scale invariance) explored in previous work and detailed by (Cael et al. 2017). The underlying reason for these shallow in-lake slopes may be related to slope-induced turbidity currents which distribute sediment from shallow high-energy areas of waterbodies to deep low-energy areas (Hakanson 1981, Johansson et al. 2007). The strength of such sediment focusing is likely greater in "younger" waterbodies with steeper slopes leading to a smoothing of their bathymetry over time (Blais and Kalff 1995).

One surprising finding with respect to the relationship between true and proxy geometry measures when examined by waterbody class was the fact that there was no greater difference between proxy and true distances in reservoirs compared to natural lakes. This is contrary to the idea that most reservoirs are drowned river valleys where the deepest point is close to the edge at the end of the reservoir (i.e. next to the dam) rather than in the center of the reservoir (Lanza and Silvey 1985). One possible explanation is that our reservoir classification data uses a more general definition of a reservoir (i.e. any permanent waterbody that has a water control structure likely to significantly impact flow or pool water) compared to that of conventional classifications that are tied to specific dam types or dam heights. Another possible explanation is that conventional reservoir classifications are conceptually biased towards more southern areas with few natural lakes (Figure 2). Southern Iowa, for instance is typically considered to have few to no natural lakes. In the present study, all of the apparent natural lakes in Southern Iowa were in fact oxbow lakes adjacent to the Missouri River.

We found other differences among waterbody geometry measures according to waterbody cross-section shape. One finding was that convex waterbodies, when compared to concave waterbodies, had longer distances to waterbody centers relative to corresponding distances to the deepest point of waterbodies. In addition, convex waterbodies often had steeper in-lake slopes relative to nearshore land slopes as compared to concave waterbodies. Finally, it was notable that convex waterbodies were deeper than concave waterbodies despite having similar distributions of waterbody surface area (Fig. S8). The underlying cause of these differences is unknown but one possibility is that geometry is tied to the circumstances of waterbody formation whereby the formation of concave waterbodies were a result of more intense glacial scouring compared to that of convex waterbodies (Gorham 1958). While our findings provide some evidence in support of this idea, namely that there is a geographic hotspot of concave waterbodies associated with the glaciated “prairie pothole region” (See Hayashi and van der Kamp 2000), the overall geographic distribution of waterbody cross-section shapes does not support this idea. Instead of a concentrated area of concave waterbodies in formerly glaciated regions, there appears to be a fairly even mix of concave and convex waterbodies distributed amongst the northern (i.e. glaciated) and southern (non-glaciated) portions of our study area (Fig. 2).

>>>>

Given the limited prediction accuracy of prior depth prediction efforts (± 6-7 m, \citealt{heathcotePredictingBathymetricFeatures2015, oliver2016prediction, sobekPredictingDepthVolume2011}), a major focus has been on improving accuracy using strategies such as employing more diverse covariates \citep{oliver2016prediction}, varying waterbody buffer sizes \citep{heathcotePredictingBathymetricFeatures2015}, or estimating hidden groupings (e.g. fitting different models for distinct size classes) among waterbodies \citep{caelVolumeMeanDepth2017, sobekPredictingDepthVolume2011}. Unfortunately, the predictive accuracy of these efforts has been limited (± 6-7 m).

One intuitive approach for predicting waterbody depth involves using a geometric model that assumes waterbody basins correspond to an idealized shape such as a cone, bowl, or an elliptic sinusoid \citep{getiranaDerivingThreeDimensional2018, hollisterPredictingMaximumLake2011, neumannMaximumDepthAverage1959, yigzawNewGlobalStorage2018}. All such geometric models for waterbody depth prediction involve implicit assumptions about the terms of geometric formulae. In the simplest case, where waterbody basins are treated as cones (Eq. \ref{eq1}, Fig. 1), two assumptions are required to make depth predictions for all waterbodies: 1) that nearshore land slope is a representative proxy for in-lake slope and 2) that the distance to the center of the waterbody is a representative proxy for the distance to the deepest point of the waterbody (Fig. 1). This cone model imposes the following fixed (i.e. geometric) relationship between slope and horizontal distance:

@@

where the product of slope and horizontal distance yields an exact geometric depth estimate ($depth_{geometric}$). Cone models of waterbody basins have been used extensively to estimate hypsography in waterbodies where there is no knowledge of volume or mean depth (Read et al. 2014, Winslow et al. 2017).

@@

The assumptions of the cone model (as well as other geometric models) can be tested by comparing proxy measures of waterbody geometry against corresponding “true” (i.e. in-lake) values derived from bathymetric maps and by evaluating how waterbody cross-section shapes differ from that of an idealized cone \citep{johanssonNewApproachesModelling2007}. For instance, waterbody cross-section shapes have been shown to vary from narrow "convex" forms to outstretched "concave" forms \citep{hakansonLakeFormLake1977}. Because tests of geometric model assumptions require bathymetric map data, which is only available for a small fraction of waterbodies (including about 15% of all waterbodies in our study footprint), existing evidence may not be applicable to all waterbodies. The few studies that have tested these assumptions have been limited to individual studies of very large (> 500 ha) waterbodies or studies on small numbers (< 100) of waterbodies \citep{johanssonNewApproachesModelling2007}. Studies focused specifically on reservoirs (as opposed to the more typical case where reservoirs and natural lakes are combined), have been even more restricted to that of large waterbodies > 1000 ha \citep{lehnerHighresolutionMappingWorld2011, messagerEstimatingVolumeAge2016}.

As a result of this limited testing, we lack knowledge on both the predictive performance of geometric models, the effect of proxies on depth prediction, and whether depth predictions are more sensitive to measurement errors in the horizontal dimension (i.e. distance to the deepest point of the waterbody) or measurement errors in the vertical dimension (i.e. in-lake slope). Additionally, it is unclear whether model prediction error is related to differences in waterbody type such those with different cross-section shapes (concave versus convex) or those classified as reservoirs versus natural lakes. To address these knowledge gaps, we asked three research questions: (1) How representative is nearshore land slope of in-lake slope; and how representative is the distance to the center of a waterbody compared to the distance to the deepest point of a waterbody? (2) How does the use of proxies for waterbody geometry affect waterbody depth prediction error? (3) How does waterbody cross-section shape (i.e. concave versus convex) and waterbody type (i.e. natural lake vs reservoir) affect depth prediction error? To answer these questions, we extracted maximum depth (hereafter referred to as “observed maximum depth”), in-lake slope, cross-section shape (i.e., concave versus convex), and distance to the deepest point, of approximately 5,000 waterbodies from bathymetric map data. We supplemented these geometry measures with data classifying waterbodies as reservoirs or natural lakes. We used this data to compute geometric depth estimates (Equation \ref{eq1}) and prediction "offsets" to these estimates using the random forest algorithm (Equation \ref{eq3}). Such offsets are model quantities which minimize differences between observed and predicted depth as a function of covariates. In our case, covariates included a variety of waterbody, watershed, and hydrologic subbasin measures that are available for all waterbodies (Table 1).

By definition, the distance proxy (distance to the center of the waterbody) must always be greater or equal to the true distance value (distance to the deepest point of the waterbody). Therefore, we expect that the use of this proxy will lead to overestimation of waterbody depth (Fig. 1). Furthermore, we expect to see greater overestimation error in reservoirs as compared to natural lakes because many reservoirs are known to be drowned river valleys where the deepest point is close to the edge at the end of the reservoir (i.e. next to the dam) rather than in the center of the reservoir (Lanza and Silvey 1985). In a similar fashion, we expect to see overestimation error associated with using a nearshore land slope proxy in waterbodies with differing cross-section shape such that the depth of bowl-shaped (i.e. concave) waterbodies will be overpredicted whereas the depth of V-shaped (i.e. convex) waterbodies will be underpredicted (Fig. S1). Finally, we expect that depth predictions themselves will be strongly related to waterbody area and hydrologic subbasin variables as these measures have been influential in prior studies \citep{oliver2016prediction}.

@@

## Methods
### Data description
We compiled bathymetry data on approximately 5,000 waterbodies in the Northeastern and Midwestern United States from nine official state databases (Fig. \ref{fig2}). These represent approximately 15% of all waterbodies in the States included in our study and are a diverse cross section in terms of their characteristics, surface areas, and span a wide geographic extent including glaciated and non-glaciated regions (Table 1). Thus, they can be considered representative of the entire population of lakes in our study extent.

@@

The original data came in a variety of formats including pre-interpolated rasters (Minnesota), contour lines (Nebraska, Michigan, Massachusetts, Kansas, Iowa), contour polygons (New Hampshire, Connecticut), or point depth soundings (Maine). For the Minnesota data, we simply clipped the raster for each waterbody to its outline. For data from the remaining States, we processed each waterbody by converting its original representation to a point layer (if necessary), rasterizing these points, and creating an interpolated bathymetry “surface” using a simple moving window average in the raster R package \citep{hijmansRasterGeographicData2019}. The size of the moving window was adjusted iteratively to ensure that each bathymetry raster contained no missing data.

All waterbody bathymetry was specifically calculated relative to high-resolution (1:24,000 scale) National Hydrography Dataset \citep{usgsNationalHydrographyDataset2019} waterbodies such that source data and bathymetry surface outputs were clipped to the area of each waterbody polygon. We restricted the waterbodies in our study to those with an area of at least 4 ha and a maximum depth of at least 0.3 m (1 ft.). The purpose of these restrictions was to ensure that waterbodies had enough contours (or points, or polygons) to generate adequately smooth interpolations with which to calculate in-lake geometry metrics.

@@

We categorized waterbodies based on their cross-section shape and reservoir class (e.g. natural lake, reservoir). For cross-section shape, we categorized waterbodies as either convex or concave following the method of \citet{hakansonLakeFormLake1977} by computing normalized waterbody depth-area relationships (i.e. hypsographic curves) and assigning class membership based on whether the midpoint of a waterbody’s curve falls above or below that of a simple straight-sided cone (Fig. S2).

We further categorized waterbodies using the output of a deep convolutional neural network model trained on satellite images labelled according to whether there was visual evidence of a water control structure significantly impacting flow \citep{polus2020}. This model had an overall validation accuracy of 81% and produced a probability for each waterbody as to whether it is a reservoir or a natural lake. For our purposes, we set a conservative classification probability threshold of 0.75 to determine whether a waterbody would be considered a reservoir. For example, if the \citet{polus2020} dataset classified a particular waterbody as a reservoir with a probability of 0.74 we categorized it as a natural lake but if the probability was greater than 0.75 we categorized it as a reservoir. Note that our reservoir classification defines reservoirs as any permanent waterbody that has a water control structure likely to significantly impact flow or pool water. It makes no distinction between different dam types, heights, or uses/purposes because the \citet{polus2020} dataset is only based on visual interpretation of waterbody images (via deep convolutional neural network models). However, the \citet{polus2020} dataset is unique in that it provides data using a standardized approach at broad spatial extents for waterbodies > 4 ha.

@@

where sinuous waterbodies have larger values of shoreline development and circular waterbodies have smaller values of shoreline development. Watershed to waterbody area ratio is an approximation of water residence time and is defined as watershed area divided by waterbody area \citep{timmsGeomorphologyLakeBasins2009}.

### Proxy evaluation
We conducted a qualitative assessment of whether or not proxy measures of waterbody geometry (e.g.  nearshore land slope, distance to the center of the waterbody) are representative of their true values (e.g. in-lake slope, distance to the deepest point of the waterbody) by visual inspection (i.e. plotting each proxy measure against its corresponding true value) and by computing coefficients of determination ($R^2$). We further tested proxy measures by examining their effect on waterbody depth prediction error. Our approach involved several steps. In the first step, we computed a geometric estimate of waterbody depth using only geometry information ($depth_{geometric}$, Equation \ref{eq1}). In the second step, we fit a random forest model to predict observed (i.e. true) depth as a function of geometric depth along with several covariates available for all waterbodies (Table 1). The purpose of this random forest “offset” modeling was to more rigorously test our expectations regarding prediction error among different formulations of $depth_{geometric}$ and among different waterbody types. Each of these steps were executed iteratively for each combination of true and proxy values of slope and distance (Table 2). We conducted additional sensitivity analysis to examine possible interactions between different proxy measures of waterbody geometry and different subsets of the entire dataset where model data was restricted (i.e. “filtered”) to include only reservoirs, only natural lakes, only convex waterbodies, or only concave waterbodies (Table 2).

## Model description
### Geometric model 
We used a geometric model of waterbodies where basins are treated as cones with a fixed relationship between slope and distance (Equation \ref{eq1}). One reason that we used the cone model is that, unlike other idealized shapes, it does not require any knowledge of waterbody volume or mean depth. Note that Equation \ref{eq1} is a geometric formula and has no intercept or coefficients and it produces an exact depth value given true values of slope and distance. To use this model to predict the depth of all waterbodies, there is a necessary assumption that proxy slope and distance measures, which are available for all waterbodies, are representative of true slope and distance (Fig. 1).

### Random forest models
Prior studies using geometric models to predict waterbody depth include a statistical or machine learning model “layer” or “offset” to boost predictive accuracy \citep{hollisterPredictingMaximumLake2011, yigzawNewGlobalStorage2018}. This procedure involves fitting a statistical or machine learning model to the residuals of an initial geometric model. For our purposes, such offset modeling enabled us to test our expectations that prediction error would be different among different formulations of $depth_{geometric}$ and among different waterbody types. It also facilitated direct comparison against prior models of waterbody depth including those that are non-geometric. We generated an “offset” to geometric depth (sensu \citealt{hollisterPredictingMaximumLake2011}) using the random forest algorithm and the ranger R package \citep{wrightRangerFastImplementation2017} to predict observed maximum depth as a function of covariates including geometric maximum depth (from Equation \ref{eq1}) along with the waterbody elevation, area, perimeter, and ratio/index measures listed in Table 2:

@@

We evaluated the relative “importance” of individual covariates by comparing model performance between 1) models where a given covariate was left untouched versus 2) models where a given covariate was permuted randomly \citep{prasadNewerClassificationRegression2006, wrightRangerFastImplementation2017}. Neither cross-section shape nor reservoir class was used as a covariate in any random forest models. Random forest training and test data was stratified on shape and reservoir class to match that of the overall waterbody population. We used the random forest algorithm because it makes no assumptions about the distribution of model residuals, allows for non-linearity, and is insensitive to interactions (i.e. multicollinearity) among covariates \citep{prasadNewerClassificationRegression2006}.

### Model comparisons
We tested model sensitivity to slope and distance proxies by generating multiple “geometric maximum depth” estimates from 3 different model runs using each of the possible metric combinations for Equation \ref{eq1} (true slope - proxy distance, proxy slope - true distance, proxy slope - proxy distance). Prior to entry into Equation \ref{eq1}, we standardized proxy distances to have the same numeric range as their true counterpart. The purpose of this standardization was to prevent waterbodies with extremely long proxy distances from having an outsized impact on model evaluation metrics. In addition to comparing among model runs using different metric combinations, we compared among sets of model runs where slope and distance measures were calculated using different sets of calculation techniques (Table S2).

### Model evaluations
We evaluated model fit and prediction error using root-mean-square error (RMSE), mean absolute percent error (MAPE), and coefficient of determination ($R^2$) metrics on a holdout set (i.e. a data subset not used for model training) containing 25% of all waterbodies. We evaluated the residuals of each model relative to waterbody cross-section shape and reservoir classes to determine whether depth is consistently over or under predicted for some waterbody types relative to others.

@@

Although proxy distance to waterbody center was often larger in magnitude compared to the true distance to the deepest point of waterbodies’ (rather than being identical), they were strongly related ($R^2$ = 0.8). Note that the coefficient of determination for this relationship is not strictly correct given that distance to waterbody center is an upper bound on distance to the deepest point of waterbodies. In contrast to distance metrics, proxy nearshore land slope and true in-lake slope were more weakly related ($R^2$ = 0.17). For slope measures, most waterbodies had higher magnitude (i.e. steeper) nearshore land slope compared to true in-lake slope (Fig. 3). Taken together, these results suggest that proxy distance to the center of waterbodies is representative of true distance to the deepest point of waterbodies whereas proxy nearshore land slope is not representative of true in-lake slope. The strong relationship between distance to the center of waterbodies and distance to the deepest point means that it is possible to convert between the two measures in subsequent analyses (See best-fit equations in Fig. 3).

@@

The use of proxy nearshore land slope had a larger effect on model fit and prediction error than the use of proxy distance to waterbody center (Table 2). More specifically, the true slope (in-lake slope) - proxy distance (distance to the center of the waterbody) model had a better fit ($R^2$ = 0.73) and lower prediction error (RMSE = 4.8m, MAPE = 27%) compared to the proxy slope - true distance model ($R^2$ = 0.31, RMSE = 7.3m, MAPE = 64%). The fit of the proxy slope -proxy distance model ($R^2$ = 0.36, RMSE = 7.1m, MAPE = 61%) was very similar to the proxy slope – true distance model. Predicted depth values for this model were generally underestimates relative to measured depth values (Fig. S9).

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.45\textwidth,keepaspectratio]{../figures/02_depth_model_importance-1.pdf}
  \end{center}
  \caption{Importance plot for random forest variables showing increase in mean square error. Higher values indicate greater importance to model predictions. See Equation \ref{eq1} for a definition of geometric max depth. HUC4 ID is a 'dummy' variable of geographic (hydrologic subbasin) location.}\label{fig5}
\end{figure}

@@

### Representativeness of proxy measures of waterbody geometry
In comparing among waterbody geometry measures, our analysis suggests that proxy distance to waterbody center is representative of true distance to the deepest point of the waterbodies, but that proxy nearshore land slope is not representative of true in-lake slope. A simple indication of this non-representativeness is that proxy nearshore land slope was often (in > 74% of cases) steeper than true in-lake slope. This finding is consistent with \citet{heathcotePredictingBathymetricFeatures2015} whose results suggest that in-lake slopes are shallower compared to the surrounding land. Furthermore, the fact that in-lake slopes were shallower compared to the surrounding land even after controlling for differences in area (Fig. S4) is consistent with the idea of topographic scaling (i.e. scale invariance) explored in previous work and detailed by \citep{caelVolumeMeanDepth2017}. The underlying reason for these shallow in-lake slopes may be related to slope-induced turbidity currents which distribute sediment from shallow high-energy areas of waterbodies to deep low-energy areas \citep{hakansonLakeBottomDynamics1981, johanssonNewApproachesModelling2007}. The strength of such sediment focusing is likely greater in "younger" waterbodies with steeper slopes leading to a smoothing of their bathymetry over time \citep{blaisInfluenceLakeMorphometry1995}.

One surprising finding with respect to the relationship between true and proxy geometry measures when examined by waterbody class was the fact that there was no greater difference between proxy and true distances in reservoirs compared to natural lakes. This is contrary to the idea that most reservoirs are drowned river valleys where the deepest point is close to the edge at the end of the reservoir (i.e. next to the dam) rather than in the center of the reservoir \citep{lanza1985interactions}. One possible explanation is that our reservoir classification data uses a more general definition of a reservoir (i.e. any permanent waterbody that has a water control structure likely to significantly impact flow or pool water) compared to that of conventional classifications that are tied to specific dam types or dam heights. Another possible explanation is that conventional reservoir classifications are conceptually biased towards more southern areas with few natural lakes (Figure 2). Southern Iowa, for instance is typically considered to have few to no natural lakes. In the present study, all of the apparent natural lakes in Southern Iowa were in fact oxbow lakes adjacent to the Missouri River.

We found other differences among waterbody geometry measures according to waterbody cross-section shape. One finding was that convex waterbodies, when compared to concave waterbodies, had longer distances to waterbody centers relative to corresponding distances to the deepest point of waterbodies. In addition, convex waterbodies often had steeper in-lake slopes relative to nearshore land slopes as compared to concave waterbodies. Finally, it was notable that convex waterbodies were deeper than concave waterbodies despite having similar distributions of waterbody surface area (Fig. S8). The underlying cause of these differences is unknown but one possibility is that geometry is tied to the circumstances of waterbody formation whereby the formation of concave waterbodies were a result of more intense glacial scouring compared to that of convex waterbodies \citep{gorhamPhysicalLimnologyNorthern1958}. While our findings provide some evidence in support of this idea, namely that there is a geographic hotspot of concave waterbodies associated with the glaciated “prairie pothole region” (see \citealt{hayashiSimpleEquationsRepresent2000}), the overall geographic distribution of waterbody cross-section shapes does not support this idea. Instead of a concentrated area of concave waterbodies in formerly glaciated regions, there appears to be a fairly even mix of concave and convex waterbodies distributed amongst the northern (i.e. glaciated) and southern (non-glaciated) portions of our study area (Fig. 2).
